The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting

### 2025-05-26 manuscript/main.tex
revision submitted

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
\mathtoolsset{showonlyrefs}
\setlength{\parindent}{0cm}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{International Journal of Production Research}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting},
  pdfkeywords={Forecasting, Pharmaceutical supply chain, Domain
knowlege, Forecast accuracy, Developing countries},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The missing puzzle piece: Contextual insights for enhanced
pharmaceutical supply chain forecasting}

@@

Given the number of products in this study, we created several data plot
and computed features of the time series, including the strength of
trend and seasonality to better understadn data.
Figure~\ref{fig-feature} shows the strength of trend versus seasonality.
Each point represents one time series with the strength of trend in
x-axis and the strength of seasonality in y-axis. Both measures are on a
scale of {[}0,1{]}. the strength of trend and seasonality were
calculated using the ``STL'' (Seasonal and Trend decomposition using
Loess) decomposition method, as described by \citet{mstl}.

@@

\subsubsection{Collaborative expert review to identify factors affecting
consumption}\label{collaborative-expert-review-to-identify-factors-affecting-consumption}

To understand the external factors influencing product consumption, we
began by visualizing the time series data for consumption as highlighted
in the previous section. This initial step helped identify specific
instances within the data that showed unusual patterns, such as peaks,
unusually high or low observations, and consecutive periods of low
consumption. These observations indicated potential events or factors
affecting consumption levels.

To gather domain knowledge about factors affected the consumption of
products, we contacted experts from EPSS who work in the distribution
unit. Next, we collaborated with six experts, each with over 10 years of
experience managing pharmaceutical distribution. In these sessions, we
reviewed the time series of all products together to understand the
potential drivers behind these anomalies. The experts provided insights
based on their extensive experience and knowledge of distribution
practices. When there was uncertainty or gaps in the data for certain
products, additional experts were consulted to gather further context.
If all experts reached a consensus, the influencing factors were
confirmed and included in the analysis.

As part of this process, we examined the electronic bin cards for 33
pharmaceutical products and conducted detailed data exploration. A bin
card is a physical or electronic record used to track the inventory
levels of individual items in a storage location, such as a warehouse or
stockroom. The term `bin' refers to the specific storage location for an
item. Information typically recorded on a bin card includes the Item ID,
unit of measure, product description, account details, date, references,
beginning balance, manufacturer, batch number, expiry date, quantity
received, quantity issued, and balance. In summary, a bin card is a
fundamental tool in inventory management, especially in environments
where physical stock tracking is essential. It provides a
straightforward and immediate way to monitor item quantities, helping to
ensure effective inventory management. Over a period of two weeks, we
meticulously traced consumption data through these bin cards. Moreover,
We ensured accuracy by converting dates from the Ethiopian calendar to
the Gregorian calendar and excluded transactions related to internal
storage or warehouse relocation within EPSS. Through this collaborative
approach, we were able to identify several predictors affecting
distribution and consumption that are summarised as followings:

\begin{itemize}
\tightlist
\item
  Stock replenishment: refers to the process of restocking or refilling
  inventory to ensure that there are sufficient quantities of products
  or materials available to meet demand. Whenever there was stock
  replenishment at the central EPSS, consumption and distribution to
  hubs and health facilities increased. This increase was attributed to
  the need to restock depleted inventories and the push from central
  EPSS to manage space constraints.
\end{itemize}

\begin{itemize}
\tightlist
\item
  Physical fiscal year inventory counting: refers to the process of
  manually counting and verifying the actual quantities of
  pharmaceutical products available in stock at a specific location. The
  process is critical for maintaining the accuracy of inventory records,
  ensuring that medicines are available when needed, and preventing
  stockouts or overstocking. Physical inventory counting periods also
  influenced consumption. Stores closed during these periods, halting
  transactions. We observed increased consumption before inventory
  counting periods, as hubs and facilities stocked up. July and August
  were identified as physical counting periods each year.
\end{itemize}

\begin{itemize}
\tightlist
\item
  Malaria seasonality: Refers to the predictable patterns and
  fluctuations in malaria incidence throughout the year, typically
  influenced by climate and environmental conditions. In many regions,
  malaria transmission peaks during and shortly after the rainy season,
  when conditions such as stagnant water pools create ideal breeding
  sites for the Anopheles mosquitoes that transmit the disease.
  Conversely, malaria cases often decline during the dry season when
  mosquito breeding sites are reduced. During peak malaria seasons,
  there is a significant surge in the demand for antimalarial drugs and
  other related treatments. Malaria seasonality was another significant
  predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine
  and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We
  identified epidemic periods affecting consumption: September to
  December 2017, March to May 2018, September to December 2018, March to
  May 2019, September to December 2019, March to May 2020, September to
  December 2020, March to May 2021, September to December 2021, and
  March to May 2022.
\end{itemize}

The key predictors identified are deterministic values, since both past
and future values of these predictors are known in advance. We then
integrated them into the consumption data, creating the complete dataset
required for model building and running the experiment.

@@

In all models, probabilistic forecasts were generated using
bootstrapping to create 1,000 possible future scenarios for each period
within the forecast horizon \citep{hyndman2021forecasting}.

@@

To assess the performance of our forecasting methods, we split the data
into a series of training and test sets and apply time series
cross-validation with a forecast horizon of 6 months. Each training set
is expanded in monthly increments, allowing model development and
hyperparameter tuning to be performed on the training data, with errors
evaluated on the corresponding test sets. We assess forecasting
performance using both point and probabilistic error measures.

@@

To measure the forecast distribution accuracy, we calculate the
Continuous Rank Probability Score \citep{hyndman2021forecasting}. It
rewards sharpness and penalizes miscalibration, so it measures overall
performance of the forecast distribution.
\begin{equation}\phantomsection\label{eq-CRPS}{
  \text{CRPS} = \text{mean}(p_j),
}\end{equation} where \[
  p_j = \int_{-\infty}^{\infty} \left(G_j(x) - F_j(x)\right)^2dx,
\] where \(G_j(x)\) is the forecasted probability distribution function
for forecast horizon \(j\), and \(F_j(x)\) is the true probability
distribution function for the same period.

@@

The forecasting performance is reported in Figure~\ref{fig-point}, in
which the average forecast accuracy over forecast horizon and across all
products is calculated for each origin. We report the distribution of
accuracy metrics across all rolling origins. This shows how models
varies in providing accuracy across different origins. The y-axis
displays models sorted by their MASE and RMSSE values, with the model
exhibiting the lowest error positioned at the bottom. This model is the
Ariam model, which incorporates exogenous variables. Additionally,
Figure~\ref{fig-point} indicates that predictors obtained through
interactions with domain experts enhance point forecast accuracy across
most models. This underscores the critical importance of systematically
collecting such expert-informed data alongside transactional consumption
data.

@@

Figure~\ref{fig-prob} illustrates the forecast distribution accuracy
measured by CRPS, which evaluates both forecasting calibration and
interval sharpness. A smaller CRPS value indicates better overall
performance. We observed that incorporating domain knowledge improved
forecast accuracy for most models, enhancing not only point forecasts
but also probabilistic forecasts. Notably, ARIMA achieved the highest
probabilistic forecast accuracy, while LSTM models had the lowest. This
aligns with the findings related to point forecast accuracy, reinforcing
the earlier explanations for these results.

We observe that LSTM models exhibit a widespread error distribution
compared to other models. This is attributed to their performance
variability, which can significantly depend on the characteristics of
demand patterns across different products. For example, the product
``Amlodipine - 5mg - Tablet'' demonstrates periods of extreme
variability, with spikes in demand followed by periods of very low or
zero demand. Such patterns align well with the strengths of LSTM models,
which excel at capturing long-term dependencies and managing complex
temporal fluctuations. In contrast, the demand for ``Anti-Rho (D)'' is
erratic and sporadic, characterized by frequent, random fluctuations
without any discernible long-term trends or seasonal patterns. This
chaotic nature, combined with inconsistent and often low volumes, poses
significant challenges for LSTM models. The lack of clear structure or
stable patterns limits the models' ability to generalize effectively,
resulting in poorer forecast performance. Although LSTM models can
handle irregular data to some extent, they require sufficient and
well-distributed training samples to achieve optimal performance. In our
study, the majority of products exhibit erratic and sporadic demand
patterns similar to ``Anti-Rho (D)'', resulting in this widespread error
distribution.

@@

Our analysis also highlights the value of using foundational time series
forecasting models like TimeGpt, particularly in developing countries
and the global health and humanitarian sectors. In these contexts, the
lack of advanced analytical skills and systematic data collection can
hinder effective forecasting and decision-making.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{todonotes,mathtools,bm,amsmath,mathpazo,xcolor}
\mathtoolsset{showonlyrefs}
\setlength{\parindent}{0cm}
\usepackage{tikz}
\usetikzlibrary{positioning, shapes.geometric, arrows.meta, fit}
\tikzset{
  process/.style={
  rectangle,
  draw=black,
  thick,
  text width=4.5cm,
  align=center,
  minimum height=1.4cm
  },
groupbox/.style={
  draw=black,
  thick,
  dashed,
  inner sep=0.3cm,
  rounded corners,
  label={[black]90:#1}
},
arrow/.style={
  thick,
  ->,
  >=Stealth
}
        }
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{International Journal of Production Research}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting},
  pdfauthor={Arebu Issa Bilal; Bahman Rostami-Tabar; Harsha Halgamuwe Hewage; Umit Sezer Bititci; Teferi Gedif Fenta},
  pdfkeywords={Forecasting, Pharmaceutical supply chain, Domain
knowlege, Forecast accuracy, Developing countries},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The missing puzzle piece: Contextual insights for enhanced
pharmaceutical supply chain forecasting}
\author[1]{Arebu Issa Bilal%
\corref{cor1}%
}
 \ead{arebu.issa@aau.edu.et} 
\author[2]{Bahman Rostami-Tabar%
%
}
 \ead{rostami-tabarb@cardiff.ac.uk} 
\author[2]{Harsha Halgamuwe Hewage%
%
}
 \ead{HalgamuweHewageHR@cardiff.ac.uk} 
\author[3]{Umit Sezer Bititci%
%
}
 \ead{U.S.Bititci@hw.ac.uk} 
\author[1]{Teferi Gedif Fenta%
%
}
 \ead{tgedif@gmail.com}

\affiliation[1]{organization={Addis Ababa University, Department of
Pharmaceutics and Social Pharmacy, School of
Pharmacy},country={Ethiopia.},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Cardiff University, Data Lab for Social
Good Reserach Group, Cardiff Business
School},country={UK.},countrysep={,},postcodesep={}}
\affiliation[3]{organization={Heriot-Watt University, Edinburgh Business
School, Heriot-Watt
University},country={UK.},countrysep={,},postcodesep={}}

@@

Given the number of products in this study, we created several data plot
and computed features of the time series, including the strength of
trend and seasonality to better understadn data.
Figure~\ref{fig-feature} shows the strength of trend versus seasonality.
Each point represents one time series with the strength of trend in
x-axis and the strength of seasonality in y-axis. Both measures are on a
scale of {[}0,1{]}. the strength of trend and seasonality were
calculated using the ``STL'' (Seasonal and Trend decomposition using
Loess) decomposition method, as described by \citet{bandara2025mstl}

@@

\caption{\label{tbl-feature}Summary statistics of the pharmaceutical
product time series}

\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{lrrrrrrrr}
\toprule
item & Mean & Median & Standard deviation & Miinum & Maximum & ACF lag 1 & zero\_run\_mean & nonzero\_squared\_cv\\
\midrule
Adrenaline (Epinephrine) - 0.1\% in 1ml Ampoule - Injection & 83077.42 & 69800.00 & 55257.59 & 0 & 204800 & 0.68 & 1.0 & 0.42\\
Amlodipine - 5mg - Tablet & 330051.67 & 84350.00 & 473283.80 & 0 & 1495700 & 0.82 & 2.0 & 1.75\\
Amoxicillin - 500mg - Capsule & 25880.97 & 25291.00 & 15751.99 & 1137 & 75305 & 0.30 & 0.0 & 0.37\\
Anti-Rho (D) & 1911.10 & 1983.00 & 1475.71 & 30 & 8477 & 0.11 & 0.0 & 0.60\\
Artemether + Lumefanthrine & 7364.28 & 953.00 & 17714.45 & 0 & 88469 & 0.64 & 6.5 & 4.33\\
Atenolol - 50mg - Tablet & 3695.75 & 3312.50 & 2587.84 & 4 & 12661 & 0.27 & 0.0 & 0.49\\
Atrovastatin - 20mg - Tablet & 3618.97 & 1302.50 & 5506.25 & 0 & 27973 & 0.44 & 1.0 & 2.26\\
Ceftriaxone & 542999.03 & 553782.50 & 346067.19 & 0 & 1285708 & 0.47 & 2.0 & 0.36\\
Dextrose & 312489.42 & 296130.00 & 255750.02 & 0 & 1004850 & 0.53 & 5.0 & 0.39\\
Frusemide - 10mg/ml in 2ml Ampoule - Injection & 176436.17 & 175360.00 & 137620.74 & 17880 & 940930 & 0.34 & 0.0 & 0.61\\
Gentamicin & 201360.50 & 182040.00 & 181410.03 & 0 & 654000 & 0.64 & 2.0 & 0.75\\
Hydralazine - 20mg/ml in 1ml Ampoule - Injection & 2952.17 & 2456.00 & 2741.23 & 6 & 15713 & 0.57 & 0.0 & 0.86\\
Insulin Isophane Human(Suspension) & 66717.83 & 73816.00 & 46364.04 & 0 & 155631 & 0.60 & 11.0 & 0.21\\
Insulin Soluble Human & 8973.33 & 8890.50 & 7140.69 & 0 & 28443 & 0.53 & 11.0 & 0.33\\
Insuline Isophane Biphasic & 12795.87 & 11402.50 & 8475.27 & 860 & 38586 & 0.45 & 0.0 & 0.44\\
Lamivudine  + Efavirenz +Tenofovir & 145178.95 & 167890.50 & 133576.98 & 0 & 550780 & 0.61 & 11.0 & 0.51\\
Lamivudine + Zidovudine & 26784.80 & 22579.00 & 22565.25 & 0 & 89910 & 0.55 & 11.0 & 0.39\\
Lidocaine HCL & 80100.70 & 62702.00 & 105600.48 & 1000 & 791050 & 0.12 & 0.0 & 1.74\\
Magnesium Sulphate & 3573.83 & 1832.50 & 4465.78 & 0 & 22649 & 0.49 & 11.0 & 1.09\\
Medroxyprogesterone & 501663.00 & 560692.00 & 328276.14 & 0 & 984774 & 0.65 & 11.0 & 0.16\\
Metformin - 500mg - Tablet & 26505.25 & 20294.75 & 21695.60 & 654 & 86705 & 0.42 & 0.0 & 0.67\\
Omeperazole - 20 mg - Capsule (Enclosing Enteric Coated Granules) & 46368.55 & 40012.00 & 38379.32 & 0 & 176219 & 0.62 & 6.0 & 0.52\\
Omeperazole - 4mg/ml in 10ml - Injection & 11438.88 & 3345.50 & 17643.52 & 0 & 81355 & 0.55 & 5.0 & 1.82\\
Oral Rehydration Salt & 507220.42 & 558426.00 & 394581.24 & 0 & 1384257 & 0.70 & 6.0 & 0.28\\
Oxytocin & 180155.23 & 151480.00 & 164309.63 & 0 & 846060 & 0.32 & 11.0 & 0.49\\
Pentavalent & 422071.73 & 474297.50 & 238940.77 & 0 & 944819 & 0.78 & 11.0 & 0.08\\
Propylthiouracil - 100mg - Tablet & 7044.46 & 6426.50 & 6112.89 & 0 & 28473 & 0.44 & 2.0 & 0.69\\
RHZ (Rifampicin + Isoniazid + Pyrazinamide) & 2260.05 & 1759.00 & 2233.00 & 0 & 7095 & 0.70 & 8.0 & 0.45\\
Rapid Diagnostic Test & 377843.65 & 360187.50 & 264551.71 & 0 & 1004325 & 0.63 & 11.0 & 0.21\\
Ringer's Injection & 88896.27 & 95227.00 & 54255.34 & 390 & 251283 & 0.53 & 0.0 & 0.37\\
Sodium Chloride (Normal Saline) & 300706.82 & 289171.50 & 136772.04 & 60759 & 702526 & 0.36 & 0.0 & 0.21\\
Sulphamethoxazole + Trimethoprim - (200mg + 40mg)/5ml - Suspension & 84216.90 & 65917.50 & 72963.22 & 20 & 259512 & 0.59 & 0.0 & 0.75\\
Tetracycline - 1\% - Eye Ointment & 199370.65 & 172721.50 & 139041.86 & 7556 & 585987 & 0.02 & 0.0 & 0.49\\
\bottomrule
\end{tabular}}

Table~\ref{tbl-feature} also provides the summary statistics computed
for each pharmaceutical product time series include traditional
distributional measures---mean, median, standard deviation, minimum, and
maximum---alongside time series--specific and structural
characteristics. These include the first-order autocorrelation (ACF at
lag 1), the mean length of consecutive zero runs (zero\_run\_mean), and
the squared coefficient of variation computed only on non-zero values
(nonzero\_squared\_cv). Together, these metrics capture not only central
tendency and dispersion, but also temporal dependence, sparsity
patterns, and relative variability in periods with non-zero consumption.

\subsection{Collaborative expert review to identify factors affecting
consumption}\label{collaborative-expert-review-to-identify-factors-affecting-consumption}

\textcolor{blue}{To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation. In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.}

Figure \ref{fig:expert-pipeline}
\textcolor{blue}{presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=0.85, transform shape, node distance=1.3cm and 2cm]

% Node style
\tikzset{
  process/.style={
    rectangle,
    draw=black,
    thick,
    text width=4.5cm,
    align=center,
    minimum height=1.4cm
  },
  arrow/.style={
    thick,
    ->,
    >=Stealth
  }
}

% Nodes
\node (step1) [process] {1. Expert Identification (with EPSS)};
\node (step2) [process, below=of step1] {2. Onboarding \& Orientation};
\node (step3) [process, below=of step2] {3. Workshop: Collaborative Data Review \& Joint Identification of Anomalies};

\node (step4) [process, right=of step1] {4. Cross Validation Using Bin Cards \& Inventory Records};
\node (step5) [process, below=of step4] {5. Structured Expert Elicitation \& Contextual Clarification};

\node (step6) [process, right=of step4] {6. Expert Elicitation of Predictor Relevance};
\node (step7) [process, below=of step6] {7. Consensus-Based Selection of Predictors};
\node (step8) [process, below=of step7] {8. Variable Encoding \& Time Alignment};

\node (step9) [process, below=of step8, yshift=-0.3cm] {9. Integration into Forecasting Models};

% Arrows
\draw [arrow] (step1) -- (step2);
\draw [arrow] (step2) -- (step3);

\path (step3.east) -- +(0.6,0) coordinate (s3stub);
\draw [arrow] (step3.east) -- (s3stub) |- (step4.west);

\draw [arrow] (step4) -- (step5);

\path (step5.east) -- +(0.6,0) coordinate (s5stub);
\draw [arrow] (step5.east) -- (s5stub) |- (step6.west);

\draw [arrow] (step6) -- (step7);
\draw [arrow] (step7) -- (step8);
\draw [arrow] (step8) -- (step9);

% Group boxes
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step1)(step2)(step3), label={[black]90:Expert Engagement}] {};
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step4)(step5), label={[black]90:Validation}] {};
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step6)(step7)(step8), label={[black]90:Feature Construction}] {};
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step9), label={[black]90:Model Integration}] {};

\end{tikzpicture}
\caption{Diagram illustrating the sequential steps involved in collecting, structuring, and utilizing domain knowledge for modeling.}
\label{fig:expert-pipeline}
\end{figure}

\textcolor{blue}{Through a facilitated half-day workshop, these experts reviewed five years of monthly consumption data for 33 pharmaceutical products. Anomalous patterns—such as consumption spikes, prolonged lows, and erratic fluctuations—were jointly identified and interpreted in light of operational events (e.g., emergency distributions, inventory cycles). These insights were cross-validated against bin cards and warehouse records, and internal transfers were excluded to isolate externally relevant events. Dates were standardized to the Gregorian calendar.}

\textcolor{blue}{Predictor variables were defined through expert consensus, with inclusion contingent on agreement among all six core participants regarding the operational relevance and causal validity of each factor. The final predictors included binary indicators for stock replenishment events, categorical variables for fiscal year inventory counts, and dummy variables for malaria seasonality. A further description of these variables are summarized as followings:}

\begin{itemize}
\item
  \textcolor{blue}{Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand. Whenever there was stock replenishment at the central EPSS, consumption and distribution to hubs and health facilities increased. This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.}
\item
  \textcolor{blue}{Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location. The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.  Physical inventory counting periods also influenced consumption. Stores closed during these periods, halting transactions. We observed increased consumption before inventory counting periods, as hubs and facilities stocked up. July and August were identified as physical counting periods each year.}
\item
  \textcolor{blue}{Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions. In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease. Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced. During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments. Malaria seasonality was another significant predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.}
\end{itemize}

\textcolor{blue}{These predictors, encoded with known historical and future values, were added to the modeling dataset. This approach ensured consistent integration of expert-informed contextual variables across both point and probabilistic forecasting models.}

@@

\subsection{Generating probabilistic
forecasts}\label{generating-probabilistic-forecasts}

\textcolor{blue}{In addition to point forecasts, we generated probabilistic forecasts to capture the uncertainty surrounding future pharmaceutical consumption. Several approaches are available for generating probabilistic forecasts, including analytical prediction intervals, quantile regression, Bayesian modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping, and conformal prediction}
\citep{wang2023}.

\textcolor{blue}{In this study, we employed a bootstrapping approach to construct predictive intervals. Bootstrapping was chosen primarily for its flexibility and model-agnostic nature, allowing it to be applied uniformly across the diverse range of forecasting methods implemented without requiring strong distributional assumptions. Moreover, pharmaceutical consumption data often exhibit irregular and volatile patterns, making non-parametric approaches like bootstrapping particularly suitable.}

\textcolor{blue}{Specifically, we assume that future forecast errors will be similar to past forecast errors. The forecast error at time $t$ is defined as: $e_t = y_t - \hat{y}_t$, where $y_t$ represents the observed consumption, and $\hat{y}_t$ denotes the corresponding forecast estimate. To simulate future consumption paths, we randomly sample errors with replacement from the historical error distribution and add them to the point forecast estimates. This process is repeated multiple times (1,000 iterations in our study) to generate a distribution of possible outcomes for each forecast horizon.}

\textcolor{blue}{Prediction intervals at the desired confidence level (e.g., 95\%) are then constructed by taking appropriate quantiles from the empirical distribution of the simulated forecasts}
\citep{hyndman2021forecasting}.

\textcolor{blue}{The bootstrapping method thus provides a robust and flexible framework to quantify forecast uncertainty across a heterogeneous set of pharmaceutical products without imposing restrictive parametric assumptions.}

@@

\textcolor{blue}{To evaluate the performance of our forecasting models, we split the data into a series of training and test sets and employed a time series cross-validation approach, following best practices for forecasting evaluation}
\citep{hyndman2021forecasting}.
\textcolor{blue}{Rather than using a fixed training and test split, we used a rolling-origin cross-validation framework, which allows for a more comprehensive assessment across different demand patterns and periods. In our setup, the initial training set consisted of all available historical data from December 2017 up to June 2021. The evaluation period covered the subsequent 12 months, reflecting the operational needs of the EPSS, which plans consumption over a one-year horizon. At each iteration, models were trained on an expanding training window and evaluated over a fixed 6-month forecast horizon, aligned with typical EPSS planning cycles. After each forecast generation, the training set was expanded by one additional month, and the process was repeated, allowing for rolling assessment across the final 12 months of data. This structure ensured that forecasts reflected realistic operational scenarios, where forecasts are continuously updated as new data becomes available. This cross-validation design allowed us to evaluate each model's ability to perform across multiple different forecast origins and a variety of demand conditions, providing a more reliable and generalizable understanding of model performance. Furthermore, all model development steps, including hyperparameter tuning for the LSTM models, were conducted exclusively using the training data available at each iteration. No test set information was used during model selection or tuning.}

@@

To measure the forecast distribution accuracy, we calculate the
Continuous Rank Probability Score \citep{hyndman2021forecasting}. It
rewards sharpness and penalizes miscalibration, so it measures overall
performance of the forecast distribution. For probabilistic evaluation,
1,000 future paths were simulated per series, enabling robust estimation
of forecast uncertainty.

\begin{equation}\phantomsection\label{eq-CRPS}{
  \text{CRPS} = \text{mean}(p_j),
}\end{equation} where \[
  p_j = \int_{-\infty}^{\infty} \left(G_j(x) - F_j(x)\right)^2dx,
\] where \(G_j(x)\) is the forecasted probability distribution function
for forecast horizon \(j\), and \(F_j(x)\) is the true probability
distribution function for the same period.

@@

The forecasting performance is reported in Figure~\ref{fig-point}, in
which the average forecast accuracy over forecast horizon and across all
products is calculated for each origin. We report the distribution of
accuracy metrics across all rolling origins. This shows how models
varies in providing accuracy across different origins. The y-axis
displays models sorted by their MASE and RMSSE values, with the model
exhibiting the lowest error positioned at the bottom. This model is the
LSTM model, followed by ARIMA, which incorporate exogenous variables.
Additionally, Figure~\ref{fig-point} indicates that predictors obtained
through interactions with domain experts enhance point forecast accuracy
across most models. This underscores the critical importance of
systematically collecting such expert-informed data alongside
transactional consumption data.

@@

Figure~\ref{fig-prob} illustrates the forecast distribution accuracy
measured by CRPS, which evaluates both forecasting calibration and
interval sharpness. A smaller CRPS value indicates better overall
performance. We observed that incorporating domain knowledge improved
forecast accuracy for most models, enhancing not only point forecasts
but also probabilistic forecasts. Notably, LSTM and ARIMA yielded the
most accurate probabilistic forecasts when identified predictors were
incorporated. This aligns with the findings related to point forecast
accuracy, reinforcing the earlier explanations for these results.

\textcolor{blue}{While LSTM models achieved the best overall forecast accuracy across products, their performance exhibited notable variability depending on the characteristics of individual demand patterns. For example, the product $Amlodipine - 5mg - Tablet$ demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero consumption. Such patterns align well with the strengths of univariate LSTM models, which are adept at capturing long-term dependencies and managing complex temporal fluctuations. In contrast, the demand for $Anti-Rho (D)$ is erratic and sparse, with frequent random fluctuations and little structural consistency. This lack of clear temporal patterns can make it challenging for LSTM models to learn generalizable signals, particularly given the limited data length available for training. Although LSTM can manage irregular data to some extent, it performs best when patterns are consistent or cyclic. These variations across product types contributed to the observed distribution of forecast errors across the product portfolio. Moreover, although we expect that multivariate LSTM models benefit from expert-informed predictors, our results show that the univariate LSTM consistently achieved better forecast performance. This outcome may be attributed to the nature of the predictors—binary, static, or weakly aligned with short-term temporal dynamics—which can disrupt rather than enhance learning when added to a neural network sensitive to input configuration. Moreover, with relatively short historical series and limited training samples, the inclusion of additional variables may have led to overfitting or reduced generalization. These findings suggest that while LSTM models can effectively learn temporal patterns from consumption data alone, incorporating structured external knowledge requires careful feature engineering and alignment to be beneficial.}

@@

\subsection{Computational efficiency and resource
considerations}\label{computational-efficiency-and-resource-considerations}

\textcolor{blue}{In addition to forecast accuracy, it is also important to consider the computational efficiency of each model, particularly in settings where computational resources and technical expertise are limited.}

Table~\ref{tbl-runtime}
\textcolor{blue}{shows the total runtime required to train and generate forecasts across all 33 pharmaceutical product time series for each method for all origins. All models were implemented using R, except TimeGPT, which was run via Google Colab using a T4 GPU backend. All other models were executed on a local machine with 7 CPU cores}
(11th Gen Intel(R) Core(TM) i5-1135G7 @ 2.40 GHz and 8 GB RAM)

\caption{\label{tbl-runtime}Computation time required for training and
generating forecasts for each model across all products.}
